Low-cost linear-scaling \emph{ab initio} molecular dynamics for weakly-interacting systems

### 2017-08-10 almo-langevin.tex
August 10
Delta Force Data Completed

\documentclass[10pt,aps,prb,twocolumn,amsmath,amssymb,superscriptaddress,longbibliography]{revtex4-1}

@@

% Re-write the following section and include any recent advances
A number of such alternative methods explore the natural sparsity of the one-electron density matrix (DM)~\cite{a:ls-kohn-1996,a:ls-rev-1999} and are capable of yielding linear scaling (LS) for large systems.~\cite{a:linscale3,a:lee-yang-1996,a:ls-scuseria-1997,a:ls-manolopoulos-1998,a:ls-helgaker-2001,a:ls-niklasson-2003,a:curvy2,a:ls-dm-sign} 
Unfortunately, the variational optimization of the DM is very inefficient for accurate DFT calculations~\cite{a:ls-rev-1999,a:ls-dm-sign}, which require many basis functions per atom. 
Therefore, the applications of DM-based LS methods have been limited to minimal-basis tight-binding problems. 
The optimal basis variants of the DM methods~\cite{a:ls-stechel-1994,a:ls-gillan-1995,a:ls-gillan-1996,a:ls-onetep-2003} designed to address this issue contract the large basis set into a small number of new localized basis functions and then optimize the DM in the contracted basis. 
Although such methods have been successfully used for the evaluation of accurate DFT energies of very large systems~\cite{a:ls-onetep-2009,a:ls-conquest-2010,a:ls-onetep-2010-app1,a:ls-rev-2012} their application for AIMD is hampered by the computationally costly optimization of both the contracted orbitals and the density matrix.
%
From this point of view, methods based on the direct optimization of localized KS orbitals~\cite{a:ls-galli-parrinello-1992,a:ls-mauri-galli-car-1993,a:ls-ordejon-1993,a:ls-mauri-galli-1994,a:ls-ordejon-1995,a:ls-kim-mauri-galli-1995,a:ls-fattebert-2004,a:ls-fattebert-2006,a:burger-yang-2008} are advantageous since they require only the occupied orbitals and, thus, deal with fewer variational degrees of freedom than the DM methods. 
Unfortunately, the progress in the development of orbital-based LS methods has been hindered by the inherently difficult convergence of the localized-orbital optimization.~\cite{a:ls-mauri-galli-car-1993,a:ls-ordejon-1995,a:ls-fattebert-2004,a:ls-rev-1999,a:weitao-yang-2013} 
%Furthermore, numerous proposed LS functionals exhibit multiple local minima, which limits their applicability in molecular dynamics simulations because of the resulting poor conservation of the total energy~\cite{a:ls-rev-1999}.

@@

This method exploits non-orthogonal molecular orbitals (MOs) in a two-step self-consistent filed (SCF) approach that is designed to avoid the usual convergence issues.~\cite{a:ls-mauri-galli-car-1993,a:ls-ordejon-1995,a:ls-fattebert-2004,a:ls-rev-1999} 
We show that due to its linear scaling nature and relatively low computational prefactor, ALMO-based DFT calculations allow for simulation of very large weakly-coupled molecular systems. 
To decrease the cost even further, the present paper builds on the ALMO SCF method as well as other recent work involving Brownian dynamics to balance force calculation errors.~\cite{a:ls-parinello,a:2ndcpmd,a:langevin-why} 
We will show that our approach lowers the computation cost pre-factor for the ALMO SCF method, while maintaining LS and accuracy for calculation of dynamical properties. 
This is done by balancing a new computationally economical but systemically inaccurate force calculation method with an optimized version of the Langevin Equation of motion. 
We also introduce an easy to follow algorithmic approach to optimize a system for generating correct dynamical behaviour.

@@

\begin{figure}
\includegraphics[trim={2.2cm 0.5cm 3cm 0.2cm},clip,width=8cm]{125mol_ALMO_Early_Random_force.eps}
\caption{\label{fig:randomforce}The average ``Random force'' along a water molecule dimension $\langle R^{\Delta}_{i\alpha} \rangle (t)$ calculated as the difference between the fully converged Born-Oppenheimer forces and unconverged ALMO DFT forces, where $\epsilon_{err} = \num{1e-2}$. 
(a) Plot of $\langle R^{\Delta}_{i\alpha} \rangle (t)$ as a function of time. 
Here it is clear that Eq.\ (\ref{eq:stochastic2}) is satisfied. 
(b) Autocovariance function of $\langle R^{\Delta}_{i\alpha} \rangle (t)$. 
The ACF is clearly not a delta-function, and is characterized by rapid oscillation between positive and negative correlation. 
However, the autocorrelation decays rapidly enough that these forces can be considered random and uncorrelated to a good approximation for sufficiently long time scales. 
Calculations were performed with the PBE $\mathrm{E_{XC}}$ functional and the TZV2P basis set on 125 water molecules.}
\end{figure}

@@

To test the accuracy of ALMO DFT based MD, we computed dynamical properties for 40 ns MD trajectories using fully converged Kohn-Sham DFT forces and forces computed with our ALMO DFT methods (see fig. %\ref{fig:dynproperties}).

>>>>

\documentclass[10pt,aps,prl,twocolumn,amsmath,amssymb,superscriptaddress,longbibliography]{revtex4-1}

@@

% Re-write the following section and include any recent advances
A number of such alternative methods explore the natural sparsity of the one-electron density matrix (DM)~\cite{a:ls-kohn-1996,a:ls-rev-1999} and are capable of yielding LS for large systems.~\cite{a:linscale3,a:lee-yang-1996,a:ls-scuseria-1997,a:ls-manolopoulos-1998,a:ls-helgaker-2001,a:ls-niklasson-2003,a:curvy2,a:ls-dm-sign} 
Unfortunately, the variational optimization of the DM is very inefficient for accurate DFT calculations,~\cite{a:ls-rev-1999,a:ls-dm-sign} which require many basis functions per atom. 
Therefore, the applications of DM-based LS methods have been limited to minimal-basis tight-binding problems. 
The optimal basis variants of the DM methods~\cite{a:ls-stechel-1994,a:ls-gillan-1995,a:ls-gillan-1996,a:ls-onetep-2003} designed to address this issue contract the large basis set into a small number of new localized basis functions and then optimize the DM in the contracted basis. 
Although such methods have been successfully used for the evaluation of accurate DFT energies of very large systems~\cite{a:ls-onetep-2009,a:ls-conquest-2010,a:ls-onetep-2010-app1,a:ls-rev-2012} their application for AIMD is hampered by the computationally costly optimization of both the contracted orbitals and the density matrix.
%
From this point of view, methods based on the direct optimization of localized KS orbitals~\cite{a:ls-galli-parrinello-1992,a:ls-mauri-galli-car-1993,a:ls-ordejon-1993,a:ls-mauri-galli-1994,a:ls-ordejon-1995,a:ls-kim-mauri-galli-1995,a:ls-fattebert-2004,a:ls-fattebert-2006,a:burger-yang-2008} are advantageous since they require only the occupied orbitals and, thus, deal with fewer variational degrees of freedom than the DM methods. 
Unfortunately, the progress in the development of orbital-based LS methods has been hindered by the inherently difficult convergence of the localized-orbital optimization.~\cite{a:ls-mauri-galli-car-1993,a:ls-ordejon-1995,a:ls-fattebert-2004,a:ls-rev-1999,a:weitao-yang-2013} 
%Furthermore, numerous proposed LS functionals exhibit multiple local minima, which limits their applicability in molecular dynamics simulations because of the resulting poor conservation of the total energy~\cite{a:ls-rev-1999}.

@@

This method exploits non-orthogonal molecular orbitals (MOs) in a two-step self-consistent filed (SCF) approach that is designed to avoid the usual convergence issues.~\cite{a:ls-mauri-galli-car-1993,a:ls-ordejon-1995,a:ls-fattebert-2004,a:ls-rev-1999} 
We show that due to its linear scaling nature and relatively low computational prefactor, ALMO-based DFT calculations allow for simulation of very large weakly-coupled molecular systems. 
To decrease the cost even further, the present paper builds on the ALMO SCF method as well as other recent work involving Brownian dynamics to balance force calculation errors.~\cite{a:ls-parinello,a:2ndcpmd,a:langevin-why} 
We will show that our approach lowers the computation cost pre-factor for the ALMO SCF method, while maintaining LS and accuracy for calculation of dynamical properties. 
This is done by balancing a new computationally economical but systemically inaccurate force calculation method with an optimized version of the Langevin Equation of motion. 
We also introduce an easy to follow heuristic approach to optimize a system for generating correct dynamical behaviour.

@@

\begin{figure}
\includegraphics[trim={0.5cm 0.7cm 0.5cm 0cm},clip,width=8cm]{DeltaForceComparison_with_ACF.eps}
\caption{\label{fig:randomforce}The average ``Random force'' along a water molecule dimension $\langle R^{\Delta}_{i\alpha} \rangle (t)$ calculated as the difference between the fully converged Born-Oppenheimer forces and ALMO DFT forces. Here it is clear that Eq.\ (\ref{eq:stochastic2}) is satisfied. 
(a) Plot of $\langle R^{\Delta}_{i\alpha} \rangle (t)$ as a function of time, where $R_{C} = 1.6$ VdW Radii and $\epsilon_{err} = \num{1e-2}$.
(b) Plot of $\langle R^{\Delta}_{i\alpha} \rangle (t)$ as a function of time, where $R_{C} = 1.6$ VdW Radii and forces are fully converged.
(c) Autocorrelation function of plot a, $\langle R^{\Delta}_{i\alpha} (t) R^{\Delta}_{i\alpha} (t+\tau) \rangle $.
Using Eq.\ \ref{eq:delta}, $\Delta = \SI{2e-5}{\per\fs}$, but its true value is $\SI{8e-5}{\per\fs}$ for this system.
Note that the  ACF is not truly a delta-function, and is characterized by rapidly oscillating correlations for short lag times. 
However, the autocorrelation decays rapidly enough that these forces can be considered random and uncorrelated to a good approximation for sufficiently long time scales. 
Calculations were performed with the PBE $\mathrm{E_{XC}}$ functional and the TZV2P basis set on 125 water molecules for 10 ps.}
\end{figure}

@@

In order to confirm that ALMO DFT forces are accurate in the limit of $R_{c} = \infty$, we computed the average difference in force between fully converged Born-Oppenheimer forces and forces calculated from ALMO DFT with varying $R_{c}$. Figure \ref{fig:forcecomp} shows that as $R_{c}$ is increased to infinity, forces converge toward the Born-Oppenheimer limit.

\begin{figure}
\includegraphics[trim={0.1cm 0cm 0cm 0cm},clip,width=8cm]{DeltaForceComparison_ALMO_SCF.eps}
\caption{\label{fig:forcecomp}The average ``Random force'' along a water molecule dimension $\langle R^{\Delta}_{i\alpha} \rangle (t)$ calculated as the average difference between fully converged Born-Oppenheimer forces and fully converged ALMO DFT forces with varying $R_{c}$.
Here it is clear that ALMO DFT forces converge to Born-Oppenheimer forces in the limit of $R_{c} = \infty$.
Calculations were performed with the PBE $\mathrm{E_{XC}}$ functional and the TZV2P basis set on 125 water molecules for 10 ps.}
\end{figure}

To test the accuracy of ALMO DFT based MD, we computed dynamical properties for 40 ns MD trajectories using fully converged Kohn-Sham DFT forces and compared with equivalent trajectories using forces computed with ALMO DFT methods (see fig. %\ref{fig:dynproperties}).
